### 2014-07-15 text/Chapter04.tex
Finish equational theory, subtyping and lambda (Chapter 4)

\subsection{Typing of let binding}

@@

Before discussing the syntactic properties of general coeffect calculus formally, it should be 
clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
equations define four syntactic reductions on the terms:
%
\begin{equation*}
\begin{array}{rllcl}
(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized substitution})\\
e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
\end{array}
\end{equation*}
%
The rules capture syntactic reductions that can be performed in a general calculus, without any 
knowledge of the specific notion of context. The \kvd{glet} notation models explicit sequencing
of context-dependent computations an is inspired by Filinski \cite{monads-inaction}. In the rest
of the section, we briefly outline the interpretation of the four rules and then we focus on
call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name (Section~\ref{sec:flat-syntax-cbn})
in more details.

@@

\paragraph{Internalized substitution.}

The (\emph{internalized substitution}) rule captures an operational semantics where the language
provides a construct representing \emph{sequential composition} and expressions can be reduced to 
a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
ordinary syntactic \kvd{let}. The normal form looks as follows:
%
\begin{equation*}
\kvd{glet}~x_1 = e_1~\kvd{in}~\ldots~\kvd{glet}~x_n = e_n~\kvd{in}~e
\end{equation*}
%
Here, the expressions $e_1, \ldots, e_n, e$ do not contain further nested \kvd{glet} constructs.
This evaluation strategy provides context-dependent counterpart to operational view of monads
developed by Filinsky \cite{monads-inaction}. We discuss how expressions reduce to the normal form
in Section~\ref{sec:flat-syntax-norm}

The (\emph{internalized substitution}) rule is useful when defining a concrete semantics for 
a language that provides constructs for explicitly providing the context. For example, consider
a language with implicit parameters where a parameter is defined by $e_1~\kvd{with}~\ident{?p}=e_2$.
Semantics for such language would provide a reduction rule:
%
\begin{equation*}
(\kvd{glet}~x_1 = \ident{?p}~\kvd{in}~e)~\kvd{with}~\ident{?p}=e_1 ~~\leadsto~~ \kvd{let}~x_1 = e_1~\kvd{in}~e
\end{equation*}
%
Here, the \kvd{glet} construct provides a way of sequentialising the context-requi\-rements so
that they can be discharged by matching constructs providing the required contexts. As discussed
tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
this form of evaluation (rather than looking at semantics for concrete systems).

@@

Recall that $\eta$-expansion turns $e$ into $\lambxa x.e~x$. In the case of liveness, the 
expression $e$ may require no variables (both immediate and latent coeffects are marked as
$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
marking the context and function argument as live. In case of data-flow, the coeffects are
made larger by the lambda abstraction. We remedy this limitation in the next chapter.

@@

\subsection{Normal form?}
\label{sec:flat-syntax-norm}

\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_2 }
  { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}

First transformation
\begin{equation*}
(\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
\end{equation*}

%
% (e2 + (e1 * e2)) + (e3 * f)
% ----(glet)------
% ----------(app)------------
%
% = e2 + e1*e2 + e3*f

% 
% (e2 + (e3 * f)) + ((e2 + (e3 * f)) * e1)
% ----(app)------    ----(app)-----
% ---------------(glet)-------------------
%
% = e2 + e3*f + e2*f + e1*e3*f

% Data-flow: (e1 + e2) max (e3 + f) = e1 + (e2 max (e3 + f))    XXXXX
% Liveness:  e2 ^ (e3 v f) = e2 ^ (e3 v f)
% Resources: OK

This does not work for dataflow, because

(e1 + e2) max (e3 + f)
e1 + (e2 max (e3 + f))

But it works for the other ones...

Second transformation
\begin{equation*}
\kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~
\kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
\end{equation*}

% e3 + ((e2 + (e1 * e2)) * e3)
%       ------(glet)---
% ---------(glet)-------------
%
% = e3 + e2*e3 + e1*e2*e3

% (e2 + (e3 * e2)) + (e1 * (e2 + (e3 * e2)))
% -----(glet)-----         -----(glet)-----
% ------------------(glet)------------------
%
% = e2 + e3*e2 + e1*e2 + e1*e2*e3

% Data-flow: e3 + (e2 + e1) = (e2 + e3) + e1
% Liveness:  e3 = e3
% Resources: OK

%
% TODO: We could have a section about let binding covering the alternative CBV and the let binding
% we included in the system, but we'd have to clean this up!
%
%\subsection{Syntactic let binding}
%\begin{remark}
%As mentioned earlier, the bounded reuse coeffect system in Section~\ref{}
%is for a CBN semantics. Recall the coeffect judgment for \emph{let} in the bounded reuse
%calculus of Section~\ref{}.
%%%
%\begin{align*}
%{\footnotesize{\trule{let}}}
%\dfrac
%{\Gamma ? \B{S} \vdash e_1 : \tau_1 \quad
%\Gamma, v : \tau_1 ? \B{R} \times T \vdash e_2 : \tau_2}
%{\Gamma ? \B{R} + (T \ast \B{S}) \vdash \synLet{v}{e_1}{e_2} : \tau_2}
%\end{align*}
%%%
%Therefore, for example, we see that \emph{let} has the coeffects
%as substitution, for example:
%%%
%\begin{align*}
%& {\footnotesize{\trule{let}}}
%\dfrac
%{y : \tau ? \langle{2}\rangle \vdash y + y : \tau \quad
%y : \tau, v : \tau ? \langle{1, 2}\rangle \vdash y + v + v : \tau_2}
%{\Gamma ? \langle{5}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2} \\
%& \hspace{3.55em} \equiv \; \Gamma ? \langle{5}\rangle \vdash y + (y + y) + (y + y) : \tau_2
%\end{align*}
%In a CBV version of bounded reuse however, the \emph{CBV let} rule gives us instead
%the judgment:
%%%
%$$
%\Gamma ? \langle{3}\rangle \vdash \synLet{v}{y + y}{y + v + v} : \tau_2
%$$
%%%
%\end{remark}
%
% 
%\paragraph{Syntactical let binding}
%Has the same typing derivation as abstraction/application
%
%\[
%\inference
% {\inference
%    { \coctx{\Gamma_1, x : \sigma}{\aclrd{R}\; \amerge \, \alift{\cclrd{t}}} \vdash e_1 : \tau }
%    { \coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{t}} \tau }
%  &
%    \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma }
%  {\coctx{\Gamma_1, \Gamma_2}{ \aclrd{R} \;\asplit\; (\cclrd{t} \,\aseq\, \aclrd{S}) } \vdash (\lambda x. e_1)~e_2 : \tau } 
%\]
%
%We can extend the CBV behaviour to \emph{let}, where the bound term
%is evaluated to a pure value first, discharging any contextual
%requirements before substitution into the body.
%
%\begin{lemma}[CBV \emph{let}]
%Let $\Gamma, x : \tau' ? \B{R} \times T \vdash e : \tau$ and 
%$\Gamma ? \B{S} \vdash e' : \tau'$. Then $\Gamma ? \B{R} \sqcup \B{S} \vdash \synLet{x}{e'}{e}
%: \tau$.
%\end{lemma}

@@

\section{Syntactic extensions}
\label{sec:flat-exts}

\subsection{Lambda abstraction}
\subsection{Constants and pairs}
\subsection{Recursion}

@@

~
\newpage
~

\newcommand{\ccat}[0]{\mathcal{C}}
\newcommand{\cobind}[2]{#1^\dagger_{#2}}
\newcommand{\cmerge}[0]{ \ident{m} }
\newcommand{\csplit}[0]{ \ident{n} }
\newcommand{\counit}[0]{ \varepsilon }
\newcommand{\llangle}{\langle\hspace{-0.25em}\langle}
\newcommand{\rrangle}{\rangle\hspace{-0.25em}\rangle}

\section{Syntax-based equational theory}
\label{sec:syntactic}

For an equational theory of coeffects, consider first a notion 
of \emph{let}-binding equivalent to $(\lambda x . e_2)~e_1$, which
has the following type and coeffect rule:
%
\begin{equation}
\inference
  {\ctyp{s}{\Gamma} \vdash e_1 : \tau_1 &
   \ctyp{r_1 \wedge r_2}{(\Gamma, x : \tau_1)} \vdash e_2 : \tau_2}
  {\ctyp{r_1 \vee (r_2 \cseq s)}{\Gamma} \vdash \kvd{let}~x = e_1~\kvd{in}~e_2
: \tau_2 }
\label{eq:let1}
\end{equation}

\noindent
For our examples, $\wedge$ is idempotent (\ie{}, $r \wedge r = r$)
implying a simpler rule:

>>>>

\subsection{Typing of let binding}
\label{sec:flat-calculus-let}

@@

Before discussing the syntactic properties of general coeffect calculus formally, it should be 
clarified what is meant by providing ``pathway to operational semantics'' in this section. We do
that by contrasting syntactic properties of coeffect systems with more familiar effect systems.
Assuming $\subst{e_1}{x}{e_2}$ is a standard capture-avoiding syntactic substitution, the following
equations define four syntactic reductions on the terms:
%
\begin{equation*}
\begin{array}{rllcl}
(\lambda x.e_1)~e_2 &{\reduce_{\ident{cbn}}}& \subst{e_1}{x}{e_2}   &&(\textit{call-by-name})\\
(\lambda x.e_1)~v   &{\reduce_{\ident{cbv}}}& \subst{e_1}{x}{v}     &\quad&(\textit{call-by-value})\\
(\lambda x.e_1)~e_2 &{\reduce_{\ident{seq}}}& \kvd{glet}~x=e_2~\kvd{in}~e_1  &&(\textit{internalized sequencing})\\
e &{\reduce_\eta}& \lambda x.e~x                                    &&(\textit{$\eta$-expansion})
\end{array}
\end{equation*}
%
The rules capture syntactic reductions that can be performed in a general calculus, without any 
knowledge of the specific notion of context. The \kvd{glet} notation models explicit sequencing
of context-dependent computations an is inspired by Filinski \cite{monads-inaction}. In the rest
of the section, we briefly outline the interpretation of the four rules and then we focus on
call-by-value (Section~\ref{sec:flat-syntax-cbv}) and call-by-name (Section~\ref{sec:flat-syntax-cbn})
in more details.

@@

\paragraph{Internalized sequencing.}

The (\emph{internalized sequencing}) rule captures an operational semantics where the language
provides a construct representing \emph{sequential composition} and expressions can be reduced to 
a normal form, consisting of a sequenced context-dependent operations. We choose to write the 
sequencing operation as \kvd{glet} to emphasize that this is a separate primitive and not an 
ordinary syntactic \kvd{let}. The normal form looks as follows:
%
\begin{equation*}
\kvd{glet}~x_1 = e_1~\kvd{in}~\ldots~\kvd{glet}~x_n = e_n~\kvd{in}~e
\end{equation*}
%
Here, the expressions $e_1, \ldots, e_n, e$ do not contain further nested \kvd{glet} constructs.
This evaluation strategy provides context-dependent counterpart to operational view of monads
developed by Filinsky \cite{monads-inaction}. We discuss how expressions reduce to the normal form
in Section~\ref{sec:flat-syntax-norm}

The (\emph{internalized sequencing}) rule is useful when defining a concrete semantics for 
a language that provides constructs for explicitly providing the context. For example, consider
a language with implicit parameters where a parameter is defined by $e_1~\kvd{with}~\ident{?p}=e_2$.
Semantics for such language would provide a reduction rule:
%
\begin{equation*}
(\kvd{glet}~x_1 = \ident{?p}~\kvd{in}~e)~\kvd{with}~\ident{?p}=e_1 ~~\leadsto~~ \kvd{let}~x_1 = e_1~\kvd{in}~e
\end{equation*}
%
Here, the \kvd{glet} construct provides a way of sequentialising the context-requi\-rements so
that they can be discharged by matching constructs providing the required contexts. As discussed
tfearlier, we focus on the general case and so we discuss when a flat coeffect calculus supports
this form of evaluation (rather than looking at semantics for concrete systems).

@@

Recall that $\eta$-expansion turns $e$ into $\lambda x.e~x$. In the case of liveness, the 
expression $e$ may require no variables (both immediate and latent coeffects are marked as
$\ident{D}$). However, the resulting expression $\lambda x.e~x$ accesses a variable, 
marking the context and function argument as live. In case of data-flow, the coeffects are
made larger by the lambda abstraction. We remedy this limitation in the next chapter.

@@

\subsection{Internalized sequencing}
\label{sec:flat-syntax-norm}

The call-by-value and call-by-name evaluation strategies discussed in the previous section are
the key techniques for defining equational theory of flat coeffects. In this section, we briefly 
discuss another approach that follows the style of generic operational semantics designed by Filinski
\cite{monads-inaction} for effectful computations.

The idea is to embed sequencing of context-dependent computations as an explicit construct into
the language and define a \emph{normal form} that consists of sequenced expressions. The reduction
to the normal form is generic for all coeffect systems (satisfying certain conditions), while the
reduction of the normal form is provided by each concrete coeffect system. The extended language
with the \kvd{glet} construct and its typing is defined as follows:
%
\begin{equation*}
\begin{array}{rcl}
e &::=& x \sep \lambda x.e \sep e_1~e_2 \sep \kvd{let}~x = e_1~\kvd{in}~e_2 \sep \kvd{glet}~x = e_1~\kvd{in}~e_2\\
n &::=& e \sep \kvd{glet}~x=e~\kvd{in}~n\\
\tau &::=& T \sep \tau_1 \xrightarrow{\cclrd{r}} \tau_2
\end{array}
\end{equation*}
\begin{equation*}
\tyrule{glet}
  { \coctx{\Gamma}{\cclrd{s}} \vdash e_1 : \tau_1 &
    \coctx{\Gamma, x\!:\!\tau_1}{\cclrd{r}} \vdash e_2 : \tau_2 }
  { \coctx{\Gamma}{\cclrd{r}\;\cpar\;({\cclrd{s}\,\cseq\,\cclrd{r})}} \vdash \kvd{glet}~x=e_1~\kvd{in}~e_2 : \tau_2 }
\end{equation*}
%
The newly introduced syntactic form $n$ represents a normal form. The construct $\kvd{glet}~x=e~\kvd{in}~n$
models an explicit sequencing of an expression $e$, followed by an expression $n$. Note that \kvd{glet} can 
only contain further \kvd{glet} constructs in the body, but not in the argument. The typing of the \kvd{glet}
is the same as the typing of ordinary \kvd{let} construct -- as discussed in Section~\ref{sec:flat-calculus-types},
for the examples we consider, this gives a \emph{more precise} typing than the typing of the
term $(\lambda x.n)~e$.

The reduction rules that produce a normal form are shown in Figure~\ref{fig:flat-normal-form}. The
(\emph{eval}) rule introduces \kvd{glet} by reducing a redex $(\lambda x.e_2)~e_1$ to an expression
representing explicit sequencing of $e_1$ and $e_2$. The remaining two rules specify how \kvd{glet}
distributes with other constructs (\kvd{glet} and application). In (\emph{glet-app}), the \kvd{glet}
construct appearing inside an application is lifted to the top-level; similarly (\emph{glet-glet}) 
lifts a \kvd{glet} construct nested in the argument of another \kvd{glet}.

As with \emph{call-by-value} and \emph{call-by-name} strategies, the \emph{internalized sequencing}
strategy can only be used with coeffect systems satisfying certain conditions. The general form of
the conditions is summarized in Appendix~?. Here, we briefly consider our three examples.

\begin{figure}[t]

\begin{equation*}
\tag{\emph{eval}}
(\lambda x.e_2)~e_1 ~~\leadsto~~ \kvd{glet}~x=e_1~\kvd{in}~e_2
\end{equation*}

\begin{equation*}
\tag{\emph{glet-app}}
\begin{array}{l}
  (\kvd{glet}~x=e_1~\kvd{in}~e_2)~e_3  ~~\leadsto~~\\ \qquad \kvd{glet}~x=e_1~\kvd{in}~(e_2~e_3)
\end{array}
\qquad x \notin \textit{fv}(e_3)\qquad
\end{equation*}

\begin{equation*}
\tag{\emph{glet-glet}}
\begin{array}{l}
  \kvd{glet}~x_2=(\kvd{glet}~x_1 = e_1~\kvd{in}~e_2)~\kvd{in}~e_3  ~~\leadsto~~\\
  \qquad \kvd{glet}~x_1=e_1~\kvd{in}~(\kvd{glet}~x_2=e_2~\kvd{in}~e_3)
\end{array}
\qquad x_1 \notin \textit{fv}(e_3)\qquad
\end{equation*}

\caption{Reduction to normal form.}
\label{fig:flat-normal-form}
\end{figure}

\paragraph{Conditions.}
The (\emph{eval}) reduction can be safely applied for any flat coeffect calculus which satisfies the
condition that the typing of the \kvd{glet} expression (shown above) is equivalent or more precise
than typing of the expression $(\lambda x.e_2)~e_1$. More formally:

\begin{definition}
A flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is \emph{oriented} if for all
$\cclrd{s}, \cclrd{s_1}, \cclrd{s_2}, \cclrd{r} \in \C$ such that $\cclrd{s}=\cclrd{s_1}\,\czip\,\cclrd{s_2}$
it is the case that $\cclrd{s} \;\cseq\; (\cclrd{s} \,\cpar\, \cclrd{r}) \;\cleq\; \cclrd{s_1} \;\cseq\; (\cclrd{s_2} \,\cpar\, \cclrd{r})$.
\end{definition}

\noindent
As discussed in Section~\ref{sec:flat-calculus-let}, this condition is satisfied for all our
three examples (strictly for liveness and data-flow; and by equality for implicit parameters).
For the (\emph{glet-app}) and (\emph{glet-glet}) rules, additional conditions arise from the
typing of the original and reduced expression (showed in Appendix~?). The results are summarized
in the following two tables.

Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_3 \xrightarrow{\cclrd{s}} \tau_2$ and
$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
reduced expressions in (\emph{glet-app}) rule are:

\begin{center}
\begin{tabular}{ | l | c | c | c |}
\hline
  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
\hspace{-1em}{\footnotesize Parameters} 
  & $\mathit{max}(\cclrd{r_1} + \cclrd{r_2}, \cclrd{r_3}+\cclrd{s})$
  & $\cclrd{r_1}+\mathit{max}(\cclrd{r_2}, \cclrd{r_3} + \cclrd{s})$ & $\times$ \\ \hline
\hspace{-1em}{\footnotesize Liveness} 
  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ 
  & $\cclrd{r_2} \sqcap (\cclrd{r_3} \sqcup \cclrd{s})$ & \checkmark \\ \hline
\hspace{-1em}{\footnotesize Data-flow} 
  & $(\cclrd{r_2} \cup \cclrd{r_1}) \cup (\cclrd{r_3} \cup \cclrd{s}) $ 
  & $(\cclrd{r_2} \cup (\cclrd{r_3} \cup \cclrd{s})) \cup \cclrd{r_1} $ & \checkmark \\ \hline
\end{tabular}
\end{center}

Assuming $\coctx{\Gamma}{\cclrd{r_1}} \vdash e_1 : \tau_1$ and
$\coctx{\Gamma}{\cclrd{r_2}} \vdash e_2 : \tau_2$ and
$\coctx{\Gamma}{\cclrd{r_3}} \vdash e_3 : \tau_3$, the typings of the original and
reduced expressions in (\emph{glet-glet}) rule are:

\begin{center}
\begin{tabular}{ | l | c | c | c |}
\hline
  & \footnotesize Before\hspace{1em} & \footnotesize After\hspace{1em}  & \footnotesize Satisfied\hspace{1em} \\ \hline
\hspace{-1em}{\footnotesize Parameters} 
  & $\cclrd{r_3} + (\cclrd{r_2} + \cclrd{r_1}) $ 
  & $(\cclrd{r_3} + \cclrd{r_2}) + \cclrd{r_1}$ & \checkmark \\ \hline
\hspace{-1em}{\footnotesize Liveness} 
  & $\cclrd{r_3}$ 
  & $\cclrd{r_3}$ & \checkmark \\ \hline
\hspace{-1em}{\footnotesize Data-flow} 
  & $\cclrd{r_3} \cup (\cclrd{r_2} \cup \cclrd{r_1}) $ 
  & $(\cclrd{r_3} \cup \cclrd{r_2}) \cup \cclrd{r_1}$ & \checkmark \\ \hline
\end{tabular}
\end{center}

This means that \emph{internalized sequencing} provides a basis for operational semantics of
liveness and implicit parameters, but not for data-flow languages. For other coeffect systems,
the conditions in Appendix~? have to be re-examined.

@@

\section{Syntactic properties and extensions}
\label{sec:flat-exts}

The flat coeffect algebra introduced in Section~\ref{sec:flat-calculus} requires a number of laws.
The laws are required for three distinct reasons -- to be able to define the categorical structure
in Section~\ref{sec:flat-semantics}, to prove equational properties in Section~\ref{sec:flat-syntax}
and finally, to guarantee intuitive syntactic properties for constructs such as 
$\lambda$-abstraction and pairs in context-aware calculi.

In this section, we look at the last point. We discuss what syntactic equivalences are permitted
by the properties of $\czip$, we extend the calculus with pairs and units and discuss their 
syntactic properties and, finall, we consider properties arising from recursion. In the following
section, we further develop subtyping relation for the calculus.

\subsection{Subtyping for coeffects}

The typing rules discussed in Section~\ref{sec:flat-calculus-types} include sub-coeffecting rule
which makes it possible to treat an expression with smaller context requirements as an expression
with greater context requirements. In the corresponding categorical semantics, this means that
we can \emph{drop} some of the provided context.

Figure~\ref{fig:flat-types-sub} adds sub-typing on function types, making it possible to treat
a function with smaller context requirements as a function with greater context requirements. 
The definition uses the standard reflexive and transitive $<:$ operator. As the (\emph{sub-fun})
shows, the function type is contra-variant in the input and co-variant in the output. The 
(\emph{typ}) rule allows using sub-typing on an expression type in the coeffect calculus.

\begin{figure}[t]

\begin{equation*}
\boxed{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau}
\end{equation*}

\begin{equation*}
\tyrule{typ}
  { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau & \tau <: \tau' }
  { \coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' }
\quad
\end{equation*}
\begin{equation*}
\tyrule{sub}
  {\coctx{\Gamma}{\cclrd{r'}} \vdash e : \tau & \cclrd{r'} \cleq \cclrd{r} }
  {\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau }
\end{equation*}

\begin{equation*}
\boxed{\tau <: \tau'}
\end{equation*}

\begin{equation*}
\tyrule{sub-trans}
  { \tau_1 <: \tau_2 & \tau_2 <: \tau_3 }
  { \tau_1 <: \tau_3  }
\end{equation*}
\begin{equation*}
\tyrule{sub-fun}
  { \tau_1' <: \tau_1 & \tau_2 <: \tau_2' & \cclrd{r'} \cgeq \cclrd{r} }
  { \tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2' }
\end{equation*}
\begin{equation*}
\tyrule{sub-refl}
  { }
  { \tau <: \tau }
\end{equation*}

\caption{Subtyping rules for flat coeffect calculus}
\label{fig:flat-types-sub}
\end{figure}

\paragraph{Semantics.} We follow the same approach as with the rest of the calculus and use
a categorical semantics to explain (and confirm) the design of the sub-typing rules. The semantics
of a judgement $\sem{\tau <: \tau'}$ is a function $\sem{\tau} \rightarrow \sem{\tau'}$. 
As shown in Figure~\ref{fig:flat-semantics-sub}, the seamntics of the sub-typing rule (\emph{typ}) 
then just composes the semantics of the original expression with the conversion produced by the 
semantics of the sub-typing judgement.

The rest of the Figure~\ref{fig:flat-semantics-sub} shows the rules that define the semantics
of $<:$. The reflexivity and transitivity are just the identity function and function composition,
respectively. The (\emph{sub-fun}) case is interesting -- recall that the semantics of a function
$\tau_1' \xrightarrow{\cclrd{r'}} \tau_2'$ is $C^{\cclrd{r'}} \tau_1' \rightarrow \tau_2'$. To 
build the required function, we first drop unnecessary context using $\ident{lift}_{\cclrd{r'}, \cclrd{r}}
 : C^{\cclrd{r'}} \tau_1' \rightarrow C^{\cclrd{r}} \tau_1'$ and use the $\ident{map}_{\cclrd{r}}$
function to transform the nested $\tau_1'$ to $\tau_1$. Then we evaluate the original function $f$
and turn the resulting $\tau_2$ into the required result of type $\tau_2'$.

\begin{figure}[t]

\begin{equation*}
\sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau' } = 
  \sem{\tau <: \tau'} \circ \sem{\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau } \qquad\quad (\emph{typ})
\end{equation*}

\begin{equation*}
\begin{array}{ll}
\sem{\tau <: \tau} = \ident{id} & (\emph{sub-refl})
\\[0.5em]
\sem{\tau_1 <: \tau_3} = \sem{\tau_2 <: \tau_3} \circ \sem{\tau_1 <: \tau_2} & (\emph{sub-trans})
\\[0.5em]
\sem{\tau_1 \xrightarrow{\cclrd{r}} \tau_2 <: \tau_1' \xrightarrow{\cclrd{r'}} \tau_2'} = \lambda f. & (\emph{sub-fun}) \\
\qquad\sem{\tau_2 <: \tau_2'} \circ f \circ \ident{map}_{\cclrd{r}}~\sem{\tau_1' <: \tau_1} 
  \circ \ident{lift}_{\cclrd{r'}, \cclrd{r}}
\end{array}
\end{equation*}

\caption{Semantics of subtyping for flat coeffects}
\label{fig:flat-semantics-sub}
\end{figure}

\subsection{Alternative lambda abstraction}

In Section~\ref{sec:flat-calculus-lambda}, we discussed how to reconcile two typings for
lambda abstraction (for implicit parameters, the lambda function splits context requirements
using $\cclrd{r} \cup \cclrd{s}$; for data-flow it suffices to duplicate the requirement 
$\cclrd{r}$). We introduced the $\czip$ operation as a way of providing the additional abstraction.
Here, we identify coeffect calculi for which the simpler (duplicating) rule is sufficient.

\paragraph{Idempotence.}
Recall that $(\C, \czip)$ is a band, meaning that $\czip$  is idempotent and associative. The 
idempotence means that the context requirements of the body can be required from both the 
declaration site and the call site. Thus, the following (\emph{idabs}) typing is valid 
(for reference, it is shown side-by-side with the ordinary lambda abstraction rule):
%
\begin{equation*}
\tyrule{idabs}
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
\tyrule{abs}
  {\coctx{\Gamma, x:\tau_1}{\cclrd{r}\,\czip\,\cclrd{r}} \vdash e : \tau_2}
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r}} \tau_2 }
\end{equation*}
% 
To derive (\emph{idabs}), we use idempotence on the body annotation $\cclrd{r}\;=\;\cclrd{r}\,\czip\,\cclrd{r}$
and then use the standard (\emph{abs}) rule. So, (\emph{idabs}) follows from (\emph{abs}), 
but the other direction is not necessarily the case. The condition below identifies 
coeffect calculi where (\emph{abs}) follows from (\emph{idabs}).

\begin{definition}
A flat coeffect algebra $(\C, \cseq, \cpar, \czip, \cunit, \czero, \cleq)$ is \emph{strictly oriented} if for all
$\cclrd{s}, \cclrd{r} \in \C$ it is the case that $\cclrd{r} \,\czip\, \cclrd{s} \;\cleq\; \cclrd{r}$.
\end{definition}

\begin{remark}
\label{thm:flat-alt-abs}
For a flat coeffect calculus with a strictly oriented algebra, the standard (abs) rule can 
be derived from the (idfun) rule.
\end{remark}
\begin{proof}
The following derives the conclusion of (\emph{abs}) using (\emph{abs}), sub-coeffecting, 
sub-typing and the fact that the algebra is \emph{stricly oriented}:

\begin{equation*}
\tyrule{typ}
  {\hspace{-5em} \tyrule{sub}
     {\hspace{-4em} \tyrule{idabs}  
        {\coctx{\Gamma, x:\tau_1}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash e : \tau_2}
        {\coctx{\Gamma}{\cclrd{r} \,\czip\, \cclrd{s}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2 } }
     {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{r} \,\czip\, \cclrd{s}} \tau_2} \;
           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}} \hspace{-5em} }
  {\coctx{\Gamma}{\cclrd{r}} \vdash \lambda x.e : \tau_1 \xrightarrow{\cclrd{s}} \tau_2} \; 
           \textnormal{\footnotesize{($\cclrd{r} \cleq \cclrd{r} \,\czip\, \cclrd{s}$)}}
\end{equation*}
\end{proof}

\noindent
The practical consequence of the Remark~\ref{thm:flat-alt-abs} is that, for strictly
oriented coeffect calculi (such as our liveness and data-flow computations), we can use
the (\emph{idabs}) rule and get an equivalent type system. This alternative formulation
removes the non-determinism that arises from the splitting of context requirements in 
the original (\emph{abs}) rule.

\paragraph{Symmetry.}
The $\czip$ operation is idempotent and associative. In all of the three examples considered in 
this chapter, the operation is also \emph{symmetric}. To make our definitions more general, we
do not require this to be the case for \emph{all} flat coeffect systems. However, systems with
symmetric $\czip$ have the following property.

\begin{remark}
For a flat coeffect calculus such that $\cclrd{r}\,\czip\,\cclrd{s} = \cclrd{s}\,\czip\,\cclrd{r}$,
assuming that $\cclrd{r'}, \cclrd{s'}, \cclrd{t'}$ is a permutation of $\cclrd{r},\cclrd{s},\cclrd{t}$:
%
\begin{equation*}
\inference
  {\coctx{\Gamma, x:\tau_1, y:\tau_2}{\cclrd{r} \,\czip\, \cclrd{s} \,\czip\, \cclrd{t}} \vdash e : \tau_3}
  {\coctx{\Gamma}{\cclrd{r'}} \vdash \lambda x.\lambda y.e : 
      \tau_1 \xrightarrow{\cclrd{s'}} (\tau_2 \xrightarrow{\cclrd{t'}} \tau_3) }
\end{equation*} 
\end{remark}

\noindent
Intuitively, this means that the context requirements of a function with multiple arguments can be 
split arbitrarily between the declaration site and (multiple) call sites. In other words, it does
not matter how the context requirements are satisfied.

\subsection{Constants and pairs}

The flat coeffect calculus annotates constants with $\czero$ (which is needed, e.g.~for liveness).
To support pairs, we add the following rule for the constructor $(e_1, e_2)$:
%
\begin{equation*}
\rulefcl{pair}
  {\coctx{r}{\Gamma} \vdash e_1 : \tau_1 & \coctx{s}{\Gamma} \vdash e_2 : \tau_2}
  {\coctx{r \fmax s}{\Gamma} \vdash (e_1, e_2) : \tau_1 \times \tau_2 }
\end{equation*}
%
The operator $\fmax$ combines context requirements of the two sub-expressions. It models
the case when the available context is split and passed to two independent expressions. The rule
for projection is uninteresting, because $\pi_i$ can be viewed as a pure function.

In a language with a constant $() : \ident{unit}$ and pairs, we would expect the following 
isomorphisms to hold:
%
\begin{equation*}
(e_1, (e_2, e_3)) \simeq ((e_1, e_2), e_3)
\quad
(e_1, e_2) \simeq (e_2, e_1)
\quad
((), e) \simeq e
\end{equation*}
%
The coeffect is not a part of the type of pairs, so the types are isomorphic. The isomorphism
of the context requirements is guaranteed by the fact that $(S, \cpar, \czero)$ is a bounded
semilattice using associativity, symmetry and boundedness, respectively.

$(\C, \cseq, \cunit)$ is monoid

$(\C, \cpar, \czero)$ is monoid

$(\C, \czip)$ is a band (associative, idempotent)

$(r \cpar s) \cseq t = (r \cseq t) \cpar (s \cseq t)$

$t \cseq (r \cpar s) = (t \cseq r) \cpar (t \cseq s)$

\subsection{Recursion}

@@
